Highly Available Transactions: Virtues and Limitations

### 2013-04-25 writeup/conclusion.tex
another pass

\section{Conclusion}
The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
However, through our analysis and experiments, we have demonstrated
that the NoSQL movement's goals of high availability will remain a
critical aspect of many future data storage systems. Accordingly, we
expose a broad design space of Highly Available Transactions (HATs),
which can offer the key benefits of highly available distributed
systems---''always on'' operation during partitions and low-latency
operations---while also providing a family of transactional isolation
levels and replica semantics that have been adopted in practice.  In
taxonomizing this design space, we show that many transactional
guarantees are achievable with high availability, while a few, like
preventing Lost Update and Write Skew, are not. From a research
perspective, this offers a long-overdue connection between storage
semantics studied in the database and distributed systems communities.

With the above understanding in hand, the next challenge is to return
to issues of performance and systems architecture within the HAT
design space.  Based on our understanding of what is desired in the
field and newfound knowledge of what is possible to achieve, we
believe HATs will be an area of great opportunity in the coming years.

>>>>

\section{Conclusion}
The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
Through our analysis and experiments, we have demonstrated how goals
of high availability will remain a critical aspect of many future data
storage systems. Accordingly, we expose a broad design space of Highly
Available Transactions (HATs), which can offer the key benefits of
highly available distributed systems---''always on'' operation during
partitions and low-latency operations---while also providing a family
of transactional isolation levels and replica semantics that have been
adopted in practice.  In taxonomizing this design space, we show that
many transactional guarantees are achievable with high availability,
while a few, like preventing Lost Update and Write Skew, are not. This
offers a long-overdue connection between storage semantics studied in
the database and distributed systems communities. The subsequent
challenge is to return to issues of performance and systems
architecture within the HAT design space.  Based on our understanding
of what is desired in the field and newfound knowledge of what is
possible to achieve, we believe HATs will be an area of great
opportunity in the coming years.

### 2013-04-25 writeup/hat-analysis.tex
another pass

HAT systems provide transactions with transactional availability and
either high availability or sticky high availability. They offer
latency and availability benefits over traditional distributed
databases, yet they cannot achieve all the traditional semantics. In
this section, we delineate ACID, distributed consistency, and session
consistency levels which can be achieved with high availability
(transactional atomicity, variants of Repeatable Read isolation, and
many session guarantees), those with sticky high availability (read
your writes, PRAM and causal consistency) and the properties which
cannot be provided in a HAT system (those preventing Lost Update and
Write Skew, or with recency).  We present a full summary of these
results in Section~\ref{sec:hat-summary}.

As Brewer states, ``systems and database communities are separate but
overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
this challenge in mind, when possible, we build on existing properties
and definitions from the database and distributed systems literature,
providing a brief, informal explanation and example for each
guarantee. The database isolation guarantees require particular care,
since different DBMSs often use the same terminology for different
mechanisms and may provide additional guarantees in addition to our
implementation-agnostic definitions.  We draw largely on Atul Adya's
dissertation~\cite{adya} and somewhat on its predecessor work: the
ANSI SQL specification~\cite{ansi-sql} and Berenson et al.'s
subsequent 1995 critique~\cite{ansicritique}. We provide a set of
formal definitions, semantics, proofs and observations in our extended
Technical Report and opt for a more informal presentation in this
paper~\cite{hat-tr}.

@@

To begin, we present well-known semantics that can be achieved in HAT
systems. We offer proof-of-concept highly available algorithms, but
our goal is \textit{not} to provide optimal or even efficient
implementations. In our examples, we exclusively consider read and
write operations. We denote a write of value $v$ to data item $d$ as
$w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$ and
assume that all data items have the null value, $\bot$, at database
initialization. Unless otherwise specified, all transactions in the
examples commit.

@@

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
$PL-1$, requiring that each transaction's writes are ordered
consistently according to a single total order on transactions. This
prohibits Adya's phenomenon $G0$, also called ``Dirty
Writes''~\cite{adya}. We form a graph of transactions, with edges from
one transaction to another if the former writes to an object before
the latter writes to the same object. PL-1 prohibits cycles in this
graph, requiring some total order on writes. Accordingly, the final
database state cannot contain the ``earlier'' transaction's writes to
the data items: for example, in the below example, $T_3$ should
\emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2, b=1$ or
$a=1, b=2$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
\end{align*}
Later properties will strengthen Read Uncommitted.

@@

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs. Implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. The implementations often provide recency and monotonicity
properties beyond the simple meaning from the name, which is what is
expressed in the implementation-agnostic definition: under Read
Committed, transactions should do not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's PL-2, and formalized by prohibiting Adya's $G1a-c$
(or ANSI's $P1$, or ``broad'' $P1$ (2.2) from Berenson et al.). For
instance, in the example below, $T_3$ should never see $a=1$, and, if
$T_2$ aborts, $T_3$ should never read $a=3$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-.5em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
transaction never writes uncommitted data to the database, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. This implementation does not provide recency or
monotonicity guarantees but satisfies the implementation-agnostic
definition.

``Repeatable Read'' isolation is a confusing property. Some IBM
products use the term for fully serializable
isolation~\cite{hat-hotos}. Gray~\cite{gray-isolation}, Berenson et
al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all interpret
Repeatable Read as providing serializability for all operations except
for predicate-based reads (and so Repeatable Read is identical to
serializability in a key-value store without predicate-based
access). In this model, the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}, is the only kind of non-serializable
behavior allowed. In this paper, we reserve the use of ``Repeatable
Read'' to refer to these definitions, which we will soon show are not
HAT-compliant.

However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
specification allows many additional behaviors under the term
Repeatable Read isolation.  \textbf{ANSI Repeatable Read} requires
that, along with respecting Read Committed isolation, each transaction
will only read one version of each data item that it did not itself
produce (preventing ``Fuzzy Read,'' or P2). In the example below,
$T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This is a literal interpretation of the phrase ``Repeatable Read'':
unless a transaction modifies a given data item, the observable value
of the data item should not change during the transaction. By itself,
this property is extremely weak---we can always read $\bot$ for each
data item---and easily achieved in a HAT system by caching the values
read.

Between our Repeatable Read (PL-2.99) and the ANSI definition, is a
concept we will call \textbf{Item Cut Isolation}: that all the
transaction's reads should see values from a non-changing consistent
cut or snapshot over the data items. It is possible to satisfy Item
Cut Isolation with high availability by using multi-versioning on each
replica and ensuring that each of a transaction's reads return an
appropriately chosen version (e.g., identified by vector clock).

A stronger achievable property requires consistent cuts to span both a
transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
its item reads.  We call this \textit{Predicate Cut
  Isolation} and it
prohibits Phantoms. Predicate Cut Isolation is also achievable in HAT
systems via similar multi-versioning as above.

@@

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property,
transactional atomicity restricts the ability to view the effects of
partially completed transactions. Under transactional atomicity, once
some of the effects of a transaction $T_1$ are observed by another
transaction $T_2$, thereafter all effects of $T_1$ are observed by
$T_2$. Together with item cut isolation, TA prevents Read Skew
anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
of TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
$b=c=1$ (or later versions for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

Perplexingly, discussions of TA are absent from existing ones of
weak isolation. This is perhaps again due to the single-node
context in which prior work was developed: on a single server (or a
fully replicated database), TA is achievable via lightweight locking
and/or local concurrency control over data items~\cite{gstore}. In
contrast, in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to know when all of a transaction's
updates are present on their respective replicas. However, replicas do
not need to agree on when to reveal a new value to clients, which
would require consensus; accordingly, TA only requires reliable
broadcast (with some additional client metadata) and is
achievable in HAT systems.

To achieve TA, clients buffer their writes until commit, sending the
final write to each data item to each respective server. Along with
each write, clients attach a the transaction ID and a list of all data
items written to in the transaction, $L_v$. Upon receiving a new write
to item $d$, a server places the write into a \textit{pending} queue
and notifies all other servers responsible for a data item in $L_v$
that it has received a write to $d$ with the respective transaction
ID. Once a server receives acknowledgments with appropriate
transaction ID for all items in $L_v$, it moves the write from
\textit{pending} to the \textit{stable} database (which applies the
writes according to highest-ID-wins). Accordingly, all writes that
should be made atomic with a write in \textit{stable} are guaranteed
to be present on their respective servers, either in \textit{pending}
or \textit{stable}. In the presence of internal aborts, writes may
require 2 RTTs: one to check constraints and one to execute the above.

For reads, each client tracks which values it should read. It
maintains a mapping (vector) from data items to transaction IDs. Upon
transaction start, the client clears its mapping, and, when it
performs a read from a data item, it sends the server its current
value for the item. If the value is empty, the server reads from
\texttt{stable}. If the value is non-empty, the server either returns
the matching value from \texttt{pending} or a higher-ID value from
\texttt{pending}---along with $L_v$ for the write. When the client
receives a response, it updates its local vector with
$L_v$. Accordingly, even if transaction's writes to two keys $x$ and
$y$ are present on the $x$ server's \textit{stable} and the $y$
server's \textit{pending}, a client's reads from $x$ then $y$ will
return a transactionally atomic pair of values because the $y$ server
will know to check pending.

This implementation is entirely masterless and both reads and writes
do not block for coordination. One key property is that we do not, in
general, dictate when writes become visible. Accordingly, it is highly
available. However, we can optimize write visibility delay by ensuring
that clients are sticky with disjoint groups of servers. Once all
transaction values are \textit{pending} within a group of servers,
values can be made \textit{stable} (as opposed to once they are
present on all servers). For example, in a multi-datacenter setting,
all clients and servers within a datacenter may form a sticky group.

@@

In the words of Berenson et al., \textit{Lost Update} occurs when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider a database containing only the following
transactions:
\begin{align*}
\small\vspace{-.5em}
T_1 &: r_x(a) w_x(a+2)
\\T_2 &: w_x(2)\vspace{-.5em}
\end{align*}
If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
operation, then the database will end up with $a=3$, a state that
could not have resulted in a serial execution due to from $T_2$'s
``Lost Update.''

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small\vspace{-.5em}
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-.5em}
\end{align*}
Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
database state could not have arisen serial execution of $T_1$ and
$T_2$.\footnote{In this example, we assume that, as is standard in
  modern databases, replicas accept values as they are written (i.e.,
  register semantics). This particular example could be made
  serializable via the use of commutative updates
  (Section~\ref{sec:evaluation}) but, as we show in the extended version of
  the paper~\cite{hat-tr}, the problem persists in the general case.}
To prevent this from happening, either $T_1$ or $T_2$ should not have
committed. Each client's respective server might try to detect that
another write occurred, but this requires knowing the version of the
latest write to $x$. In our example, this reduces to a requirement for
linearizability, which is, via Gilbert and Lynch's proof of the CAP
Theorem, provably unachievable with high
availability~\cite{gilbert-cap}.

\textbf{Write Skew} is a generalization of Lost Update to multiple
keys. It occurs when one transaction $T1$ reads a given data item $x$,
a second transaction $T2$ reads a different data item $y$, then $T1$
writes to $y$ and commits and $T2$ writes to $x$ and commits. As an
example of Write Skew, consider the following two transactions:
\begin{align*}
\small
T_1 &: r_y(0)~w_x(1)
\\T_2 &: r_x(0)~w_y(1)
\end{align*}
As Berenson et al. describe, if there was an integrity constraint
between $x$ and $y$ such that only one of $x$ or $y$ should have value
$1$ at any given time, then this write skew would violate the constraint (which is preserved in serializable executions). Write skew is a somewhat
esoteric anomaly---for example, it does not appear in
TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
Update, it is also unavailable to HATs.

@@

Data storage systems make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g.,
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force an available system to violate any recency bound, so recency
bounds are not enforceable by HAT systems.

@@

We summarize our results in Table~\ref{table:hatcompared}. A wide
range of isolation levels are achievable in HAT systems, including
transactional atomicity, cut isolation, and several session
guarantees. With sticky availability, a system can achieve read your
writes guarantees and PRAM and causal consistency. However, many other
prominent models, such as Snapshot Isolation, One-Copy
Serializability, and Strict Serializability cannot be achieved due to
the inability to prevent Lost Update and Write Skew phenomena.

We illustrate the hierarchy of available, sticky available, and
unavailable consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable, but we find several particularly compelling. If we combine
all sticky-HAT guarantees, we have transactional, causal snapshot
reads (i.e., Causal Transactional Predicate Cut Isolation). If we
combine TA and P-CI, we have transactional snapshot reads. We can
achieve RC, MR, and RYW by simply sticking clients to servers. And we
can also combine unavailable models---for example, an unavailable
system might provide PRAM and One-Copy
Serializability~\cite{daudjee-session}.

@@

\subsection{Discussion}
\label{sec:discussion}

In this section, we discuss several subtleties in our results,
specifically addressing model composition, transactional atomicity
versus linearizability, and stickiness requirements.

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. Transactional atomicity
dictates that writes to multiple keys across multiple servers are made
visible to readers all at once, while linearizability dictates that
writes to a single key on multiple servers are made visible to all
readers at once---what is different? First, in linearizable (and safe
and regular) systems, writes are made visible to all clients
\textit{immediately} after they finish. With transactional atomicity,
there is no recency guarantee. Second, in linearizabile systems, all
clients see all writes at the same time. With transactional atomicity,
clients may see writes at different times depending on which replicas
they contact. We are not aware of an analogous model in the
distributed systems literature. Accordingly, despite apparent
similarities, transactional atomicity is incomparable with and much
cheaper (by availability standards) than linearizability.

\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
availability can result in much better write \textit{visibility}:
clients will be able to safely read writes more quickly in a sticky
available system. In the model we discussed, it is possible to achieve
several properties like monotonic reads in a highly available system
by waiting to reveal a write until all servers have seen it and its
relevant dependencies. However, this incurs severe visibility
penalties---new writes will not become visible to clients in the
presence of partitions. A client that does not want to guarantee
read-your-writes (due to the sticky availability requirement) may
still wish to read other clients' writes with timeliness.

>>>>

HAT systems provide transactions with transactional availability and
either high availability or sticky high availability. They offer
latency and availability benefits over traditional distributed
databases, yet they cannot achieve all possible semantics. In this
section, we delineate ACID, distributed consistency, and session
consistency levels which can be achieved with high availability
(transactional atomicity, variants of Repeatable Read isolation, and
many session guarantees), those with sticky high availability (read
your writes, PRAM and causal consistency), and properties that cannot
be provided in a HAT system (those preventing Lost Update and Write
Skew, or with recency).  We present a full summary of these results in
Section~\ref{sec:hat-summary}.

As Brewer states, ``systems and database communities are separate but
overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
this challenge in mind, when possible, we build on existing properties
and definitions from the database and distributed systems literature,
providing a brief, informal explanation and example for each
guarantee. The database isolation guarantees require particular care,
since different DBMSs often use the same terminology for different
mechanisms and may provide additional guarantees in addition to our
implementation-agnostic definitions.  We draw largely on Atul Adya's
dissertation~\cite{adya} and somewhat on its predecessor work: the
ANSI SQL specification~\cite{ansi-sql} and Berenson et al.'s
subsequent 1995 critique~\cite{ansicritique}. We provide a set of
formal definitions, semantics, proofs and observations in our extended
Technical Report~\cite{hat-tr} and opt for a more informal
presentation here.

@@

To begin, we present well-known semantics that can be achieved in HAT
systems. We offer proof-of-concept highly available algorithms, but
our goal is \textit{not} to provide optimal or even efficient
implementations. As is standard in the distributed database
literature, in our sample implementations, we consider a set of
servers acting as replicas for a set of data items with which
non-co-located clients interact. In our examples, we exclusively
consider read and write operations, denoting a write of value $v$ to
data item $d$ as $w_d(v)$ and a read from data item $d$ returning $v$
as $r_d(v)$. We assume that all data items have the null value,
$\bot$, at database initialization, and, unless otherwise specified,
all transactions in the examples commit.

@@

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}, requiring that each transaction's writes are ordered
consistently according to a single total order on transactions. This
prohibits Adya's phenomenon $G0$, also called ``Dirty
Writes''~\cite{adya}. More formally, we require a total ordering on
writes to each data item and we create a graph of transactions with
edges from one transaction to another if the former overwrites the
latter's write to the same object. \textit{PL-1} prohibits cycles in
this graph, requiring some total order on writes. Accordingly, the
final database state cannot contain the ``earlier'' transaction's
writes to the data items: for example, in the below example, $T_3$
should \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2,
b=1$ or $a=1, b=2$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
\end{align*}
Later properties will strengthen Read Uncommitted.

@@

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs. Implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. The implementations often provide recency and monotonicity
properties beyond the simple meaning from the name, which is what is
expressed in the implementation-agnostic definition: under Read
Committed, transactions should do not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's \textit{PL-2} and is formalized by prohibiting
Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
from Berenson et al.). For instance, in the example below, $T_3$
should never see $a=1$, and, if $T_2$ aborts, $T_3$ should never read
$a=3$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-.5em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of the databaseq, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. This implementation does not provide recency or
monotonicity guarantees but satisfies the implementation-agnostic
definition.

``Repeatable Read'' isolation is a confusing property. For example,
some IBM products use the term for fully serializable
isolation~\cite{hat-hotos}, while Gray~\cite{gray-isolation}, Berenson
et al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except for predicate-based reads (and so Repeatable Read is
identical to serializability in a key-value store without
predicate-based access). In this latter model, the \textit{Phantom
  Problem}, whereby two successive predicate-based reads return
different data~\cite{gray-isolation}, is the only kind of
non-serializable behavior allowed. In this paper, we reserve the use
of ``Repeatable Read'' to refer to these definitions, which we will
soon show are not HAT-compliant.

However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
specification allows several additional behaviors under ``Repeatable
Read'' isolation.  \textbf{ANSI Repeatable Read} requires that, along
with respecting Read Committed isolation, each transaction will only
read one version of each data item that it did not itself produce
(preventing ``Fuzzy Read,'' or P2). In the example below, $T_3$ must
read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This is a literal interpretation of the phrase ``Repeatable Read'':
unless a transaction modifies a given data item, the observable value
of the data item should not change during the transaction. By itself,
this property is weak---we can always read $\bot$ for each data
item---and easily achieved in a HAT system by caching the values
read. To prevent confusion, we call this isolation model \textbf{Item
  Cut Isolation}: all the transaction's reads should see values from a
non-changing consistent cut or snapshot over the data items. It is
possible to satisfy Item Cut Isolation with high availability by using
multi-versioning on each replica and ensuring that each of a
transaction's reads return an appropriately chosen version (e.g.,
identified by vector clock).

A stronger achievable property that expands on Item Cut Isolation
requires consistent cuts to span both a transaction's predicate reads
(e.g., \texttt{SELECT WHERE}) as well as its item reads.  We call this
\textbf{Predicate Cut Isolation} and it prohibits Phantoms. Predicate
Cut Isolation is also achievable in HAT systems via similar
multi-versioning as above.

@@

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property, TA
restricts transactions' ability to view the effects of partially
completed transactions. Under TA, once some of the effects of a
transaction $T_i$ are observed by another transaction $T_j$,
thereafter all effects of $T_i$ are observed by $T_2$. Together with
item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
A5A~\cite{ansicritique}). As an example of TA, because $T_2$ has read
$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

Perplexingly, discussions of TA are absent from existing treatment of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), TA is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore}. In contrast,
in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to ensure that, regardless of replica
selection, all of a client's accesses will read a sufficiently
up-to-date/``consistent'' set of data items. However, replicas do not
need to agree on when to choose to reveal new sets of values clients,
which would require consensus. Instead, TA only requires the
equivalent of reliable broadcast (with some additional client
metadata) and is achievable in HAT systems.

We can achieve TA with a two-phase multi-versioning algorithm. Clients
buffer their writes until commit, sending the final write to each data
item to each respective server. Along with each write, clients attach
a the transaction ID and a list of all data items written to in the
transaction, $L_v$. Upon receiving a new write to item $d$, a server
places the write into a \texttt{pending} queue and notifies all other
servers responsible for a data item in $L_v$ that it has received a
write to $d$ with the respective transaction ID. Once a server
receives acknowledgments with appropriate transaction ID for all items
in $L_v$, it moves the write from \texttt{pending} to the
\texttt{stable} database (which applies the writes according to
highest-ID-wins). Accordingly, all writes that should be made atomic
with a write in \texttt{stable} are guaranteed to be present on their
respective servers, either in \texttt{pending} or
\texttt{stable}. Given the possibility of internal aborts, writes to
\texttt{pending} may require 2 RTTs: one to check constraints and one
to execute the above (but these checks \textit{cannot} cause other
transactions to block, as would be the case in a two-phase
\textit{commit} protocol).

For reads, each client tracks which values it should read. It
maintains a mapping (vector) from data items to transaction IDs. Upon
transaction start, the client clears its mapping, and, when it
performs a read from a data item, it sends the server its current ID
for the item. If the client does not supply an ID, the server returns
the latest value from \texttt{stable}. If the ID is non-empty, the
server either returns the matching value from \texttt{pending} or a
value with higher ID from \texttt{pending}---along with $L_v$ for the
write. When the client receives a response, it updates its local
vector with $L_v$. Accordingly, even if transaction's writes to two
keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
and the replica for $y$'s \texttt{pending}, a client's reads from $x$
then $y$ will return a transactionally atomic pair of values because
the replica for $y$ will know to check \texttt{pending}.

This implementation is entirely masterless and both reads and writes
do not block for coordination. One key property is that we do not, in
general, dictate when writes become visible. Accordingly, it is highly
available. However, we can optimize write visibility delay by ensuring
that clients are sticky with disjoint groups of servers. Once all
transaction values are \texttt{pending} within a group of servers,
values can be made \texttt{stable} (as opposed to once they are
present on all servers). For example, in a multi-datacenter setting,
all clients and servers within a datacenter may form a sticky group.

@@

Berenson et al. define \textit{Lost Update} as when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider a database containing only the following
transactions:
\begin{align*}
\small\vspace{-.5em}
T_1 &: r_x(a) w_x(a+2)
\\T_2 &: w_x(2)\vspace{-.5em}
\end{align*}
If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
operation, then the database will end up with $a=3$, a state that
could not have resulted in a serial execution due to from $T_2$'s
``Lost Update.''

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small\vspace{-.5em}
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-.5em}
\end{align*}
Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
database state could not have arisen serial execution of $T_1$ and
$T_2$.\footnote{In this example, we assume that, as is standard in
  modern databases, replicas accept values as they are written (i.e.,
  register semantics). This particular example could be made
  serializable via the use of commutative updates
  (Section~\ref{sec:evaluation}) but the problem persists in the
  general case.}  To prevent this from happening, either $T_1$ or
$T_2$ should not have committed. Each client's respective server might
try to detect that another write occurred, but this requires knowing
the version of the latest write to $x$. In our example, this reduces
to a requirement for linearizability, which is, via Gilbert and
Lynch's proof of the CAP Theorem, provably unachievable with high
availability~\cite{gilbert-cap}.

\textbf{Write Skew} is a generalization of Lost Update to multiple
keys. It occurs when one transaction $T1$ reads a given data item $x$,
a second transaction $T2$ reads a different data item $y$, then $T1$
writes to $y$ and commits and $T2$ writes to $x$ and commits. As an
example of Write Skew, consider the following two transactions:
\begin{align*}
\small
T_1 &: r_y(0)~w_x(1)
\\T_2 &: r_x(0)~w_y(1)
\end{align*}
As Berenson et al. describe, if there was an integrity constraint
between $x$ and $y$ such that only one of $x$ or $y$ should have value
$1$ at any given time, then this write skew would violate the constraint (which is preserved in serializable executions). Write skew is a somewhat
esoteric anomaly---for example, it does not appear in
TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
Update, it is also unavailable to HAT systems.

@@

Distributed data storage systems often make various recency guarantees
on reads of data items. As we have discussed, one of the most famous
is linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g.,
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force an available system to violate any recency bound, so recency
bounds are not enforceable by HAT systems.

@@

As we summarize in Table~\ref{table:hatcompared}, a wide range of
isolation levels are achievable in HAT systems, including
transactional atomicity, cut isolation, and several session
guarantees. With sticky availability, a system can achieve read your
writes guarantees and PRAM and causal consistency. However, many other
prominent models, such as Snapshot Isolation, One-Copy
Serializability, and Strict Serializability cannot be achieved due to
the inability to prevent Lost Update and Write Skew phenomena.

We illustrate the hierarchy of available, sticky available, and
unavailable consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable, but we find several particularly compelling. If we combine
all sticky-HAT guarantees, we have transactional, causal snapshot
reads (i.e., Causal Transactional Predicate Cut Isolation). If we
combine TA and P-CI, we have transactional snapshot reads. We can
achieve RC, MR, and RYW by simply sticking clients to servers. We can
also combine unavailable models---for example, an unavailable system
might provide PRAM and One-Copy
Serializability~\cite{daudjee-session}.

@@

\subsection{Additional Discussion}
\label{sec:discussion}

In this section, we discuss several subtleties in our results,
specifically addressing model composition, transactional atomicity
versus linearizability, and benefits of stickiness.

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. Transactional atomicity
dictates that writes to multiple keys across multiple servers are made
visible to readers all at once, while linearizability dictates that
writes to a single key on multiple servers are made visible to all
readers at once---what is different? First, in linearizable (and safe
and regular) systems, writes are made visible to clients
\textit{immediately} after they finish. With transactional atomicity,
there is no recency guarantee. Second, in linearizabile systems, all
clients see all writes at the same time. With transactional atomicity,
clients may see writes at different times depending on which replicas
they contact. We are not aware of an analogous model in the
distributed systems literature. Accordingly, despite apparent
similarities, transactional atomicity is incomparable with and much
cheaper (by availability standards) than linearizability.

\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
availability can improve write \textit{visibility}: clients will be
able to safely read writes more quickly in a sticky available
system. In the model we discussed, it is possible to achieve several
properties like monotonic reads in a highly available system by
waiting to reveal a write until all servers have seen it and its
relevant dependencies. However, this incurs severe visibility
penalties---new writes will not become visible to clients in the
presence of partitions. A client that does not want to guarantee
read-your-writes (due to the sticky availability requirement) may
still wish to read other clients' writes with timeliness.
